---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# vrtility

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of vrtility is to make the best use of GDAL's VRT capabilities for
efficient processing of large raster datasets. This package's primary focus is 
on the use of GDAL VRT pixel functions using [numba](https://numba.pydata.org/).
for very efficient raster processing. At present the only function provided
is a simple median. There is definitely scope to expand this functionality! 
Comments and contributions are most welcome! 

## Features

- No intermediate downloads - the use of nested VRTs enables the download
and processing of only the required data in a single gdalwarp call. 

- use of numba in python pixel function(s)

## TO DO:

- [ ] Add efficient masking.
- [ ] Add more pixel functions (geometric median in particular).
- [ ] clean things up a lot!
- [ ] time series functions...
- [ ] configure GPU processing with numba?

...

## Installation

You can install the development version of vrtility from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Permian-Global-Research/vrtility")
# next set up the required Python environment
vrtility::build_vrtility_python()
```

## Example

Here is a simple example where we: define a bounding box, search a STAC catalog
for Sentinel-2 data, create a VRT, and then create a composite image 
using a median pixel function.

```{r example}
library(vrtility)
library(tictoc)

bbox <- gdalraster::bbox_from_wkt(
  wkt = wk::wkt("POINT (130.75 -11.45)"),
  extend_x = 0.1, extend_y = 0.125
)

s2_vrt <- sentinel2_stac_vrt(
  bbox = bbox,
  start_date = "2023-01-01",
  end_date = "2023-04-30",
  assets = c(
    # "B01",
    "B02", "B03", "B04"
    # "B05", "B06", "B07",
    # "B08", "B8A", "B09",
    # "B11", "B12"
    # , "SCL"
  )
)

print(s2_vrt)

tic()
s2_composite <- vrt_composite(
  s2_vrt,
  outfile = fs::file_temp(ext = ".tif"),
  bbox = bbox
)
toc()

terra::plotRGB(
  terra::rast(s2_composite), 3, 2, 1,
  stretch = "lin", zlim = c(700, 2200), zcol = TRUE,
  mar = c(2, 2, 2, 2),
  axes = TRUE
)
```

