---
title: "Harmonized Landsat Sentinel-2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Harmonized Landsat Sentinel-2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```





```{r setup}
# library(vrtility)
devtools::load_all()

mirai::daemons(6)

bbox <- gdalraster::bbox_from_wkt(
  wkt = wk::wkt("POINT (144.3 -7.6)"),
  extend_x = 0.17,
  extend_y = 0.125
)

te <- bbox_to_projected(bbox)
trs <- attr(te, "wkt")
```


```{r hlssl}
hlssl_stac <- hls_stac_query(
  bbox = bbox,
  start_date = "2023-01-01",
  end_date = "2023-12-31",
  max_cloud_cover = 35,
  collection = "HLSL30_2.0"
)
print(hlssl_stac)
```

```{r hlsss}
hlsss_stac <- hls_stac_query(
  bbox = bbox,
  start_date = "2023-01-01",
  end_date = "2023-12-31",
  max_cloud_cover = 35,
  collection = "HLSS30_2.0"
)
print(hlsss_stac)
```

```{r earthdata-creds}
earthdatalogin::edl_netrc(
  username = Sys.getenv("EARTHDATA_USER"),
  password = Sys.getenv("EARTHDATA_PASSWORD")
)
```

```{r hls-collect}
hls_sl_col <- vrt_collect(hlssl_stac)
print(hls_sl_col)

hls_ss_col <- vrt_collect(hlsss_stac)
print(hls_ss_col)
```

```{r hls-mask}
hls_sl_col_mask <- vrt_set_maskfun(
  hls_sl_col,
  mask_band = "Fmask",
  mask_values = c(0, 1, 2, 3),
  build_mask_pixfun = build_bitmask(),
  drop_mask_band = TRUE
)

hls_ss_col_mask <- vrt_set_maskfun(
  hls_ss_col,
  mask_band = "Fmask",
  mask_values = c(0, 1, 2, 3),
  build_mask_pixfun = build_bitmask(),
  drop_mask_band = TRUE
)

# let's check out a vrt and inspect the mask function that we used.
print(hls_sl_col_mask, maskfun = TRUE)
```


HLS data for the landsat and sentinel collections contain different numbers
of bands. They share X bands but both have sensor-specific bands. Confusingly,
the two collections also have shared names that represent different bands and
wavelengths. So, let's rename all the bands so that every wavelength is uniquely
named and then add empty bands to each collection so that the number of bands is the
same for each collection.

```{r band-align-hls}
hls_ls_align <- vrt_set_band_names(
  hls_sl_col_mask,
  c("A", "B", "G", "R", "N2", "S1", "S2", "C", "T1", "T2")
) |>
  vrt_add_empty_band(after = 4, description = "RE1") |>
  vrt_add_empty_band(after = 5, description = "RE2") |>
  vrt_add_empty_band(after = 6, description = "RE3") |>
  vrt_add_empty_band(after = 7, description = "N") |>
  vrt_add_empty_band(after = 9, description = "WV") |>
  vrt_move_band(band_idx = 13, after = 10)



hls_ss_align <- vrt_set_band_names(
  hls_ss_col_mask,
  c("A", "B", "G", "R", "RE1", "RE2", "RE3", "N", "N2", "WV", "C", "S1", "S2")
) |>
  vrt_add_empty_band(after = 13, description = "T1", scale_value = 0.01) |>
  vrt_add_empty_band(after = 14, description = "T2", scale_value = 0.01)
```

Now we can combine the two collections into a single collection.

```{r hls-merge}
hls_merge_coll <- c(
  hls_ls_align,
  hls_ss_align
)
```

Now we need to warp all vrt_blocks in the collection to the same spatial
reference system, extent and pixel size.

```{r hls-warp}
hls_warp <- vrt_warp(
  hls_merge_coll,
  t_srs = trs,
  te = te,
  tr = c(30, 30),
  resampling = "bilinear"
)
```

Now we can stack the warped vrt collection - this can be thought of as a virtual 
raster cube.

```{r hls-stack}
hls_stack <- vrt_stack(hls_warp) |>
  vrt_set_pixelfun(pixfun = median_numpy())
```

```{r hls-compute}
outfile <- tempfile(fileext = ".tif")
hls_stack <- vrt_compute(hls_stack, outfile, engine = "gdalraster", nsplits = 3)
```